### 2017-04-04 paper.tex
Add glossary

\input{header.tex}

@@

@@

\begin{abstract}

\end{abstract}

\newpage
\tableofcontents
\setlength{\parskip}{1em}

\newpage
\chapter{Introduction}
\label{ch:introduction}

@@

Most HAB missions led by universities, schools or amateur are built around
commercially available small computers or micro-controller boards: Redland
Green School students used a Raspberry Pi (\cite{rpi2014}) running
python scripts as an On Board Computer (\textit{OBC}) for their HAB mission
(\cite{Hinschelwood2015}), while other projects have been ran on AVR
micro-controller based boards like the Arduino Mega
(\cite{AtmelCorporation2015}).

In the United Kingdom, radio licenses do not cover airborne transmission, making
packet radio protocols like AX.25 unusable for HAB missions 
(\cite{ukhasradio2016}). However, the Office of Communications allows the use
of certain \textit{license-exempt} frequencies like the 70cm wavelength band,
provided that the transmission power does not exceed 10mW (\cite{Ofcom2014}).
For this reason, most missions launched from the United Kingdom use the
Radiometrix NTX2 transmitter (\cite{radiometrix2012}). Because of the
constraints, these missions usually transmit their telemetry as text using the
Radio-Teletype (RTTY) protocol.

While most HAB projects make use of commercial-off-the-shelf components, the
finished payloads are usually custom, single-purpose designs that are little,
if at all modular. In the design document for the Titan-1 HAB mission, Bombasaro
describes a hardware bus based on tightly coupled sensors which communicate
over two different data buses, SPI and I2C, and a custom-made, single-purpose
flight software (\cite{Bombasaro2015}).

Because of the risks and cost involved, nano-satellites rely more on modular
designs, allowing each team to work on their module independently. In his paper,
Volstad describes the design of the data bus of the NTNU CubeSat: while the OBC
is a based on a custom made circuit board, it is designed to provide a standard
power interface, as well as access to an I2C standard bus
(\cite{NXPSemiconductors2014}) to each payload (\cite{Volstad2011}). The I2C
protocol was chosen because of its low power consumption, and because if only
requires two lines (clock and data). Thomas Clausen describes in his 2001 paper
how a simple packet protocol built on top of I2C itself is used for data
transfer and error detection in Aalborg University's CubeSats
(\cite{Clausen2001}).

@@

There are however some technologies that can be adapted from satellites to be
used in HAB applications. The University of Arizona uses a custom packet format
over 434MHz radio, containing raw binary values for each instrument's
measurements (\cite{Eatchel2002}). The BRITE-Austria CubeSat mission uses the
AX.25 packet radio protocol (used by amateur radio users), which allows the
use of off-the-shelf transmission and reception hardware rather than custom-made
circuits (\cite{Traussnig2007}).

Some nano-satellites use custom telemetry data format, which allows the team
to minimise the volume of data to be sent over radio. The Planetary Society's
Lightsail mission has an uplink connection that allows the ground station
to request specific logs or data. By default, The spacecraft only communicates
a custom beacon containing a summary of the its housekeeping data
(\cite{planetary2016}). Uplink being impractical with the transmission power
limits imposed in the United Kingdom, such a selective telemetry system cannot
be relied upon for HAB missions.

@@

SpaceWire defines how instruments' data can be accessed by having the OBC
remotely poll their memory (\cite{parkes2005}). A similar system is used to
fetch data from ROM chips over I2C and could be used to build a simple data bus.

The CCSDS Space Packet protocol defines a packet format used
to encapsulate application data (from instruments for example) that can be sent
over a ``Space Link'', an analogue of the OSI model's link layer
(\cite{Stallings1987}) using CCSDS frames to encapsulate packets originating
from multiple instruments, ground stations and spacecrafts (\cite{ccsds2003}).

While HAB applications do not require the level of complexity of CCSDS
protocols — addressing is not needed since the network only contains two
endpoints, communicating in a single direction — some details of the Space
Packet protocol are worth reusing (Forward Error Correction, different
Application IDs for each payload).

@@

The error-correction requirement led to the use of a Forward Error Correction
(\textit{FEC}) algorithm. FEC algorithms function by adding some amount of data
\(D\) to each message sent \(M\). \(D\) is computed by the encoder. The decoder
uses \(D\) to detect whether errors were introduced anywhere in the received
block \(M+D\), and correct those errors if their number is below the
algorithm's maximum.

@@

\item Reed-Solomon codes are transparent: the original data is not modified for
the purposes of error correction. This means that even if too many errors are
present for the FEC decoder to correct, the data can be manually inspected and
some of it can potentially be saved.

@@

Figure \ref{fig:frame-fsm} presents the finite state machine that AHABus frame
decoders must implement to detect frames in a binary stream, and figure
\ref{fig:frame-fmt-3} the final version of the AHABus frame format.

@@

\begin{figure}[H]
    \begin{bytefield}{16}
        \bitheader{0,7,8,15} \\
        \begin{leftwordgroup}{Header}
            \bitbox{8}{\texttt{0x5A}} & \bitbox{8}{version} \\
            \bitbox{16}{sequence number}
        \end{leftwordgroup} \\
        \begin{leftwordgroup}{Payload}
            \wordbox[lrt]{1}{data} \\
            \skippedwords \\
            \wordbox[lrb]{1}{}
        \end{leftwordgroup} \\
        \begin{leftwordgroup}{FEC}
            \wordbox[lrt]{1}{checksum} \\
            \skippedwords \\
            \wordbox[lrb]{1}{}
        \end{leftwordgroup}
    \end{bytefield}
    \centering
    \caption{AHABus Frame Format - Version 3}
    \label{fig:frame-fmt-3}
\end{figure}

@@

Encapsulating the transmitted data adds overhead. If \(n\) is the size of a
chunk in bytes, and each frame adds \(m\) bytes of metadata, then only a
percentage \(\frac{n}{n+m}\) of the stream contains actual data. For reasons
explained further, the AHABus frames were chosen to be 256-byte long. As shown
previously, each frame also includes.

@@

@@

\newpage
\printbibliography[heading=bibnumbered, title=References]

>>>>

\input{header.tex}
\input{glossaries.tex}

@@

\frontmatter

@@

\begin{abstract}
\end{abstract}

\tableofcontents
\setlength{\parskip}{1em}

\listoffigures

\listoftables

\printglossary
\printglossary[type=\acronymtype]

\mainmatter

\chapter{Introduction}
\label{ch:introduction}

@@

Most \acrshort{hab} missions led by universities, schools or amateur are built 
around commercially available small computers or micro-controller boards:
Redland Green School students used a Raspberry Pi (\cite{rpi2014}) running
python scripts as an \acrfull{obc} for their HAB mission
(\cite{Hinschelwood2015}), while other projects have been ran on AVR
micro-controller based boards like the Arduino Mega
(\cite{AtmelCorporation2015}).

In the United Kingdom, radio licenses do not cover airborne transmission, making
packet radio protocols like AX.25 unusable for HAB missions 
(\cite{ukhasradio2016}). However, the Office of Communications allows the use
of certain \textit{license-exempt} frequencies like the 70cm wavelength band,
provided that the transmission power does not exceed 10mW (\cite{Ofcom2014}).
For this reason, most missions launched from the United Kingdom use the
Radiometrix NTX2 transmitter (\cite{radiometrix2012}). Because of the
constraints, these missions usually transmit their telemetry as text using the
\acrfull{rtty} protocol.

While most \acrshort{hab} projects make use of commercial-off-the-shelf
components, the finished payloads are usually custom, single-purpose designs
that are little, if at all modular. In the design document for the Titan-1 HAB
mission, Bombasaro describes a hardware bus based on tightly coupled sensors
which communicate over two different data buses, SPI and \acrshort{i2c}, and a
custom-made, single-purpose flight software (\cite{Bombasaro2015}).

Because of the risks and cost involved, nano-satellites rely more on modular
designs, allowing each team to work on their module independently. In his paper,
Volstad describes the design of the data bus of the NTNU CubeSat: while the OBC
is a based on a custom made circuit board, it is designed to provide a standard
power interface, as well as access to an \acrshort{i2c} standard bus
(\cite{NXPSemiconductors2014}) to each payload (\cite{Volstad2011}). The
\acrshort{i2c} protocol was chosen because of its low power consumption, and
because if only requires two lines (clock and data). Thomas Clausen describes
in his 2001 paper how a simple packet protocol built on top of \acrshort{i2c}
itself is used for data transfer and error detection in Aalborg University's
CubeSats (\cite{Clausen2001}).

@@

There are however some technologies that can be adapted from satellites to be
used in \acrshort{hab} applications. The University of Arizona uses a custom
packet format over 434MHz radio, containing raw binary values for each
instrument's measurements (\cite{Eatchel2002}). The BRITE-Austria CubeSat
mission uses the AX.25 packet radio protocol (used by amateur radio users),
which allows the use of off-the-shelf transmission and reception hardware
rather than custom-made circuits (\cite{Traussnig2007}).

Some nano-satellites use custom telemetry data format, which allows the team
to minimise the volume of data to be sent over radio. The Planetary Society's
Lightsail mission has an uplink connection that allows the ground station
to request specific logs or data. By default, The spacecraft only communicates
a custom beacon containing a summary of the its housekeeping data
(\cite{planetary2016}). Uplink being impractical with the transmission power
limits imposed in the United Kingdom, such a selective telemetry system cannot
be relied upon for \acrshort{hab} missions.

@@

SpaceWire defines how instruments' data can be accessed by having the OBC
remotely poll their memory (\cite{parkes2005}). A similar system is used to
fetch data from ROM chips over \acrshort{i2c} and could be used to build a
simple data bus.

The \acrshort{ccsds} Space Packet protocol defines a packet format used to
encapsulate application data (from instruments for example) that can be sent
over a ``Space Link'', an analogue of the OSI model's link layer
(\cite{Stallings1987}) using \acrshort{ccsds} frames to encapsulate packets
originating from multiple instruments, ground stations and spacecrafts
(\cite{ccsds2003}).

While \acrshort{hab} applications do not require the level of complexity of
\acrshort{ccsds} protocols — addressing is not needed since the network only
contains two endpoints, communicating in a single direction — some details of
the Space Packet protocol are worth reusing (Forward Error Correction,
different Application IDs for each payload).

@@

The error-correction requirement led to the use of a Forward Error Correction
(\textit{FEC}) algorithm. FEC algorithms function by adding some amount of data
\(D\) to each message sent \(M\). In this case, the message includes both the
frame's header and its payload. \(D\) is computed by the encoder. The decoder
uses \(D\) to detect whether errors were introduced anywhere in the received
block \(M+D\), and correct those errors if their number is below the
algorithm's maximum.

@@

\item Reed-Solomon codes are transparent: the original data is not modified for
the purposes of error correction. This means that even if too many errors are
present for the \acrshort{fec} decoder to correct, the data can be manually
inspected and some of it can potentially be saved.

@@

Figure \ref{fig:frame-fsm} presents the finite state machine that
\acrshort{ahabus} frame decoders must implement to detect frames in a binary
stream, and figure \ref{fig:frame-fmt-3} the final version of the
\acrshort{ahabus} frame format.

@@

\begin{figure}[H]
    \begin{bytefield}{16}
        \bitheader{0,7,8,15} \\
        \begin{leftwordgroup}{Header}
            \bitbox{8}{\texttt{0x5A}} & \bitbox{8}{version} \\
            \bitbox{16}{sequence number}
        \end{leftwordgroup} \\
        \begin{leftwordgroup}{Payload}
            \wordbox{2}{data[]}
        \end{leftwordgroup} \\
        \begin{leftwordgroup}{FEC}
            \wordbox{2}{checksum[32]}
        \end{leftwordgroup}
    \end{bytefield}
    \centering
    \caption{\acrshort{ahabus} Frame Format - Version 3}
    \label{fig:frame-fmt-3}
\end{figure}

@@

Encapsulating the transmitted data adds overhead. If \(n\) is the size of a
chunk in bytes, and each frame adds \(m\) bytes of metadata, then only a
percentage \(\frac{n}{n+m}\) of the stream contains actual data. For reasons
explained further, the \acrshort{ahabus} frames were chosen to be 256-byte
long. As shown previously, each frame also includes.

@@

To evaluate the protocol's viability, a reference implementation of an encoder
and decoder was written. The goal was to

@@

\printbibliography[heading=bibintoc, title=References]
